Challenges on quantifying fish range shifts across management boundaries

### 2021-08-26 Scripts/manuscript_acrossboundaries.Rmd
included authors

---
title: "Challenges on quantifying fish range shifts across management boundaries"
subtitle: "DRAFT. Target journal *PlosOne*"
output: 
  bookdown::word_document2:
    number_section: no
    toc: no
---

@@

Range shifts are among the most documented responses of marine fauna to the effects of climate change in the ocean [@Poloczanska]. From zooplankton to bony fish to seabirds, marine fauna have been documented to expand their distribution up to 800 km per decade or to contract it for about 200 km per decade [@Poloczanska]. Such change is often due to marine species following a poleward temperature gradient in the search for colder waters, although, in some cases such species are following local environmental factors such as upwelling systems [@Clarke:2020], or moving to deeper colder waters [@Pinsky:2013]. Regardless of the shift direction, species distributions are projected to continue to happen in the incoming years [@Cheung:2010; @Barange2014; @IPCC2019], even if society engages in a strong mitigation climate change scenario (i.e., limiting increased global temperature to $1.5^o$) as proposed by the Paris Agreement [@Cheung:2016].

@@

In the United States (US) North East (NE) Atlantic, many important fished species such as american lobster (*Homarus americanus*), yellowtail flounder (*Limanda ferruginea*), summer flounder (*Paralichthys dentatus*) and red hake (*Urophycis chuss*) have been shifting their distribution poleward  [@pinsky_fogarty_2012], and modelling exercises suggest that NE Atlantic species will shift will average over 600 km towards the end of the $21^{st}$ century under a high emission climate change scenario [@Morley:2018]. The mismatch between implemented management strategies and shifts in stocks across management jurisdictions compromises management and sustainability [@Baudron2020; @pinsky_fogarty_2012]. Fishers in the NE Atlantic have been adapting to such shifts by following the fish, in some cases, or changing target speices, in others [@polo_2021]. While this problem of shifting stocks across management jurisdictions is increasingly well known, quantifying it is difficult for two main reasons: (*i*) management jurisdiction in the ocean are often ill-defined, especially at the sub-national level or beyond Exclusive Economic Zones (EEZ), and (*ii*) quantifying fish distributions in relation to these boundaries is difficult, even where fishery-independent surveys exist, because of seasonal and interannual variation.

Here, we examine both of these challenges by quantifying the historical distributions of three important fished species relative to 11 states of the Northeast Atlantic U.S that jointly manage these stocks. Specifically, we set two main objectives. First, we examine the difference between two different approaches to define spatial management jurisdictions within the U.S. Federal waters. Second, we examine differences in the species' historical distribution between Spring and Fall to examine seasonal variation in these stock's distributions. We use bottom trawl surveys for Spring and Fall from 1971 to 2019 and apply a triangular irregular surface method to extrapolate fish biomass across the U.S. North East EEZ. We then draw spatial polygons for each State up to the EEZ from two different starting points; the state waters and main fishing ports. Our results suggest that...

@@

Biomass data for all three species was gathered from NOAA's yearly Northeast Fisheries Science Center Spring and Fall Bottom Trawl Surveys [data](https://www.fisheries.noaa.gov/region/new-england-mid-atlantic#science) provided by [Ocean adapt](https://oceanadapt.rutgers.edu/) and standardized as catch per unit of effort (CPUE) [@Pinsky2013]. The analysis was retrospective in nature, covering the time period from 1971 to 2019 with two comparative periods; the average of 1980 to 2000 representing the state allocations of the commercial fisheries originally implemented in 2003 [@RefDocument], and the average of 2009 to 2019 representing a more recent time period.

We extrapolated fish CPUE observational data to the NE Atlantic EEZ from the southern border of North Carolina to the US-Canadian border by following a triangular irregular surface (TIS) method using the *R* package *interp* [@REFInterp]. Overall, the TIS method creates multiple non-overlapping triangles that cover the entire extrapolated region (i.e., the EEZ) from an initial set of sample points (i.e., NOAAs survey) generating a spatially-continuous surface [@REF; @REF].

Federal U.S. waters are not sub-dived into state-owned fisheries management regions. Thus, we created state-polygons to allocate the extrapolated CPUE to each fishing State. This step allowed us to further determine what proportion of the species distribution was in each state-polygon. The state-polygons were created by first expanding multiple polygons within the 200 nautical miles of the EEZ using the *st* package in *R* [@REF]. Second, we gridded the EEZ into 3 x 3 gridcells and overlay the grid with the state-polygons to then estimate the number of grids that each expanded polygon occupied. Note that a gridcell would be allocated to multiple States in cases where state-polygons overlay. Finally, the buffers were limited to the size and shape of the North East US EEZ.

Creating these state-polygons was key to allocate the distribution of any stock directly influencing the proportion of the stock linked to a state. For example, the amount of the distribution that each State has will be influenced by the size of the state-polygon. Here, we used two different sources as starting points for the state-polygons expansion; state waters and US ports (Figure \@ref(fig:bufferFig)). For the state waters approach we used the [NOAA-data.gov](https://catalog.data.gov/dataset/federal-and-state-waters) shapefile. Note that in this case, the extent of the final state-polygons was related to the size of the State's coast with North Carolina and Maine having the largest polygons. For the ports approach we used the [Global Fishing Watch](https://globalfishingwatch.org/datasets-and-code-anchorages/)'s database on Anchorages, Ports and Voyages. We only considered ports classified within the World Port Index ([WPI](https://msi.nga.mil/NGAPortal/MSI.portal?_nfpb=true&_pageLabel=msi_portal_page_62&pubCode=0015)). In this case, the number of ports varied among States with New Jersey and Massachusetts having the largest amounts (117 and 187, respectively) and Delaware and New Hampshire having the leaser (2 and 14, respectively). However, rather than the number of ports per State, in this case what influences the size of the state-polygon was the distribution of ports within the State's waters as two ports in opposite latitudes will result in a larger polygon than two ports closed together (Figure \@ref(fig:bufferFig)).

@@

For each approach we then overlay each specie's interpolation with the state-polygons and estimated the distribution and biomass proportion of each state. The interpolated biomass was divided equally in cases where two or more States had the same grid. For example, if New York and New Jersey both shared one grid, then the proportion of that grid would be allocated 50% to each state, alternatively, if Connecticut was also sharing the grid, then the proportion would be 33.3% to each state. Note that this step does not affect the interpolation method but the way the biomass is allocated by State.

@@

# Call the image
img <- readPNG("/Volumes/Enterprise/Data/AcrossBoundaries/Results/Figures/F2_proportion_chng_agg")
# Print the image
grid.raster(img)
```

![Figure 2. Change in the proportion of the distribution today (2014-2019) relative to 1973-1985. A) row-wise showing seasonal changes, column-wise showing changes by spatial approach. B) Differnces between seasons for each spatial approach. C) Differences between spatial approaches for each season.](/Volumes/Enterprise/Data/AcrossBoundaries/Results/Figures/F2_proportion_chng_agg)

@@

Changes will be different when looked at the species level (Figure \@ref{fig:figSppHist}). All three species show a reduction in biomass proportion for southern states (namely Virginia and North Carolina) with black sea bass and scup having the larger declines. Biomass increases are less clear. For black sea bass, biomass has increased from Maine all the way down to Rhode Island in the Fishing ports approach, and New York for the State waters approach. In the State waters approach the species has just recently "arrived" to Maine and New Hampshire waters (Figure \@ref{fig:figSppHist}- State waters). For scup, only Maine and New Hampshire seem to have gain biomass following the Fishing ports approach while the State waters approach resulted in gains for all States northern than New York. Similar to black sea bass, Virginia and North Carolina are the main biomass losers in both approaches. Finally, there does not appear to be clear evidence of shifts in biomass for summer flounder. Decreases are seen for Virginia and North Carolina for both approaches and increases for Maine and New Hampshire for the fishing ports approach, yet, these are almost imperceptible (Figure \@ref{fig:figSppHist}).

>>>>

---
title: "Challenges on quantifying fish range shifts across management boundaries"
subtitle: "DRAFT. Target journal *PlosOne*"
author: "Palacios-Abrantes Juliano, Crosson Scott, Dumas Chris, Fujita Rod, Jensen Olaf, Levine Arielle, Longo Catherine"
output: 
  bookdown::word_document2:
    number_section: no
    toc: no
---

@@

Range shifts are among the most documented responses of marine fauna to the effects of climate change in the ocean [@Poloczanska]. From zooplankton to bony fish to seabirds, marine fauna has been documented to expand their distribution up to 800 km per decade or to contract it for about 200 km per decade [@Poloczanska]. Such change is often due to marine species following a poleward temperature gradient in the search for colder waters, although, in some cases such species are following local environmental factors such as upwelling systems [@Clarke:2020], or moving to deeper colder waters [@Pinsky:2013]. Regardless of the shift direction, species distributions are projected to continue to happen in the incoming years [@Cheung:2010; @Barange2014; @IPCC2019], even if society engages in a strong mitigation climate change scenario (i.e., limiting increased global temperature to $1.5^o$) as proposed by the Paris Agreement [@Cheung:2016].

@@

In the United States (US) North East (NE) Atlantic, many important fished species such as american lobster (*Homarus americanus*), yellowtail flounder (*Limanda ferruginea*), summer flounder (*Paralichthys dentatus*) and red hake (*Urophycis chuss*) have been shifting their distribution poleward  [@pinsky_fogarty_2012], and modelling exercises suggest that NE Atlantic species will shift will average over 600 km towards the end of the $21^{st}$ century under a high emission climate change scenario [@Morley:2018]. The mismatch between implemented management strategies and shifts in stocks across management jurisdictions compromises management and sustainability [@Baudron2020; @pinsky_fogarty_2012]. Fishers in the NE Atlantic have been adapting to such shifts by following the fish, in some cases, or changing target speices, in others [@polo_2021]. While this problem of shifting stocks across management jurisdictions is increasingly well known, quantifying it is difficult for two main reasons: (*i*) management jurisdictions in the ocean are often ill-defined, especially at the sub-national level or beyond Exclusive Economic Zones (EEZ), and (*ii*) quantifying fish distributions in relation to these boundaries is difficult, even where fishery-independent surveys exist, because of seasonal and interannual variation.

Here, we examine both of these challenges by quantifying the historical distributions of three important fished species relative to 11 states of the Northeast Atlantic U.S that jointly manage these stocks. Specifically, we set two main objectives. First, we examine the difference between two different approaches to define spatial management jurisdictions within the U.S. Federal waters. Second, we examine differences in the species' historical distribution between Spring and Fall to examine seasonal variation in these stock's distributions. We use bottom trawl surveys for Spring and Fall from 1971 to 2019 and apply a triangular irregular surface method to extrapolate fish biomass across the U.S. Northeast EEZ. We then draw spatial polygons for each State up to the EEZ from two different starting points; the state waters and main fishing ports. Our results suggest that...

@@

Biomass data for all three species was gathered from NOAA's yearly Northeast Fisheries Science Center Spring and Fall Bottom Trawl Surveys [data](https://www.fisheries.noaa.gov/region/new-england-mid-atlantic#science) provided by [Ocean adapt](https://oceanadapt.rutgers.edu/) and standardized as catch per unit of effort (CPUE) [@Pinsky2013].

with two comparative periods; the average of 1980 to 2000 representing the state allocations of the commercial fisheries originally implemented in 2003 [@RefDocument], and the average of 2009 to 2019 representing a more recent time period.

We extrapolated fish CPUE observational data to the NE Atlantic EEZ from the southern border of North Carolina to the US-Canadian border by following a triangular irregular surface (TIS) method using the *R* package *interp* [@REFInterp]. Overall, the TIS method creates multiple non-overlapping triangles that cover the entire extrapolated region (i.e., the EEZ) from an initial set of sample points (i.e., NOAAs survey) generating a spatially-continuous surface [@REF; @REF]. The extrapolation was done separately  for both Fall and Spring seasons covering the time period of 1971 to 2019.

Federal U.S. waters are not sub-dived into state-owned fisheries management regions. Thus, we created state-polygons to allocate the extrapolated CPUE to each fishing State. This step allowed us to further determine what proportion of the species distribution was in each state-polygon. The state-polygons were created by first expanding multiple polygons within the 200 nautical miles of the EEZ using the *st* package in *R* [@REF]. Creating these state-polygons was key to allocate the distribution of any stock directly influencing the proportion of the stock linked to a state. For example, the amount of the distribution that each State has will be influenced by the size of the state-polygon. Here, we used two different sources as starting points for the state-polygons expansion; state waters and US ports (Figure \@ref(fig:bufferFig)). For the state waters approach we used the [NOAA-data.gov](https://catalog.data.gov/dataset/federal-and-state-waters) shapefile. Note that in this case, the extent of the final state-polygons was related to the size of the State's coast with North Carolina and Maine having the largest polygons. For the ports approach we used the [Global Fishing Watch](https://globalfishingwatch.org/datasets-and-code-anchorages/)'s database on Anchorages, Ports and Voyages. We only considered ports classified within the World Port Index ([WPI](https://msi.nga.mil/NGAPortal/MSI.portal?_nfpb=true&_pageLabel=msi_portal_page_62&pubCode=0015)). In this case, the number of ports varied among States with New Jersey and Massachusetts having the largest amounts (117 and 187, respectively) and Delaware and New Hampshire having the leaser (2 and 14, respectively). However, rather than the number of ports per State, in this case what influences the size of the state-polygon was the distribution of ports within the State's waters as two ports in opposite latitudes will result in a larger polygon than two ports closed together (Figure \@ref(fig:bufferFig)).

@@

We then gridded the EEZ into 3 x 3 grid-cells and overlay it with the state-polygons to identify the grids that each state-polygon occupied. Note that a grid-cell would be allocated to multiple States in cases where state-polygons overlay. For each approach we then overlay each specie's interpolation with the state-polygons and estimated the distribution and biomass proportion of each state. The interpolated biomass was divided equally in cases where two or more States had the same grid. For example, if New York and New Jersey both shared one grid, then the proportion of that grid would be allocated 50% to each state, alternatively, if Connecticut was also sharing the grid, then the proportion would be 33.3% to each state. Note that this step does not affect the interpolation method but the way the biomass is allocated by State.

Finally, we estimated the biomass percentage change today relative to the historic period. For that matter we define the historic period as 1980 to 2000 representing the state allocations of the commercial fisheries originally implemented in 2003 [@RefDocument], and “today” as the average of 2009 to 2019 representing a more recent data available. The parentage changes between the historic and present biomass ($\Delta{B}$) was estimate for each of the four approaches as;

$$ \Delta B=\frac{(B_t-B_h)}{|B_h|}*100$$

Where $B_t$ is the biomass today (average of 2009 to 2019) and $B_h$ is the biomass at the historical period (average of 1980 to 2000). In addition, we estimated the differences in ($\Delta{B}$) between the seasonal ($\Delta{B}_{seas}$) and the spatial ($\Delta{B}_{spat}$) approach by subtracting from each other:

$$ \Delta B_{seas} = |\Delta B_{fp}-\Delta B_{sw}| $$
$$ \Delta B_{spat} = |\Delta B_{f}-\Delta B_{s}| $$

Where $\Delta{B}_{fp}$, $\Delta{B}_{sw}$, $\Delta{B}_{f}$, and $\Delta{B}_{s}$ represent the biomass parentage change today relative to historical value for fishing ports, state waters, Fall and Summer approaches, respectively.

@@

# Call the image
img <- readPNG("/Volumes/Enterprise/Data/AcrossBoundaries/Results/Figures/F2_proportion_chng_agg.png")
# Print the image
grid.raster(img)
```

![Figure 2. Change in the proportion of the distribution today (2014-2019) relative to 1973-1985. A) row-wise showing seasonal changes, column-wise showing changes by spatial approach. B) Differnces between seasons for each spatial approach. C) Differences between spatial approaches for each season.](/Volumes/Enterprise/Data/AcrossBoundaries/Results/Figures/F2_proportion_chng_agg.png)

@@

All three species show a reduction in biomass proportion for southern states (namely Virginia and North Carolina) with black sea bass and scup having the larger declines (Figure \@ref{fig:figSppHist}). Biomass increases are less clear. For black sea bass, biomass has increased from Maine all the way down to Rhode Island in the Fishing ports approach, and New York for the State waters approach. In the State waters approach the species has just recently "arrived" to Maine and New Hampshire waters (Figure \@ref{fig:figSppHist}- State waters). For scup, only Maine and New Hampshire seem to have gain biomass following the Fishing ports approach while the State waters approach resulted in gains for all States northern than New York. Similar to black sea bass, Virginia and North Carolina are the main biomass losers in both approaches. Finally, there does not appear to be clear evidence of shifts in biomass for summer flounder. Decreases are seen for Virginia and North Carolina for both approaches and increases for Maine and New Hampshire for the fishing ports approach, yet, these are almost imperceptible (Figure \@ref{fig:figSppHist}).
